Cultural Transmission Can Explain the Evolution of Cooperation

### 2020-07-29 ms/main.tex
finished revising result 2 (formerly 1)

Therefore, for cooperation to take over the population (for $\tilde{p}=1$ to be globally stable) we require $\tilde{p}'>\tilde{p}$, that is,
\begin{equation}
  \tilde{p}^2 (1+b-c) + \tilde{p}(1-\tilde{p}) \big[(1-c) (1 - T_B) + (1+b)T_A\big] 
  > \bar{w}\tilde{p} .
\end{equation}
We divide by $\tilde{p}$, set $\bar{w} = 1 + \tilde{p}(b-c)$, and rearrange to get
\begin{equation}
  (1-\tilde{p}) \big[(1-c) (1 - T_B) + (1+b)T_A\big] 
  > 1 -\tilde{p} .
\end{equation}
Dividing by $(1-\tilde{p})$ we find that $\tilde{p}'>\tilde{p}$ if 
\begin{equation} \label{eq:vert_hori_alpha1_global_condition}
  (1-c) (1 - T_B) + (1+b)T_A
  > 1 \Leftrightarrow
  c < \frac{bT_A + (T_A- T_B)}{1-T_B}.
\end{equation}
\\

@@

We define the initial frequency as $\tilde{p}_0$ and
\begin{equation}\begin{aligned}
\gamma_1 = \frac{b \alpha T_A - (T_A - T_B)}{1-T_B}, \quad
\gamma_2 = \frac{b \alpha T_B + (1+b) (T_A - T_B)}{1-T_B}.
\end{aligned}\end{equation}
Applying eqs.~\ref{eq:vert_hori_equilibrium}, \ref{eq:vert_hori_global_condition1}, and \ref{eq:vert_hori_global_condition2} we can summarize the possible outcomes:
\begin{enumerate}
\item Fixation of cooperation,
	if $T_A>T_B$ and $c < \gamma_1$, or
	if $T_A>T_B$ and $\gamma_1 < c < \gamma_2$, and $\tilde{p}_0>\tilde{p}^*$, or
	if $T_A<T_B$ and $\gamma_2 < c$.
\item Global fixation of defection,
	if $T_A>T_B$ and $\gamma_2 < c$, or
	if $T_A<T_B$ and $c < \gamma_1$, or
	if $T_A<T_B$ and $\gamma_1 < c < \gamma_2$, and $\tilde{p}_0<\tilde{p}^*$.
\item Co-existence of both phenotypes at $\tilde{p}^*$, 
	if $T_A<T_B$ and
	$\gamma_2 < c < \gamma_1$.
\end{enumerate}

Importantly, the condition for cooperation to increase from rarity ($\tilde{p}_0 \approx 0$) is
\begin{equation} \label{eq:unequal_transmission_from_rarity}
T_A>T_B \; \text{and} \; c < \gamma_1
\end{equation} 
Thus, cooperation must have a horizontal transmission bias, and the cost of cooperation cannot be too big.

@@

\begin{corollary}[No assortment of transmission and cooperation]
When $\alpha=0$, 
if there is a horizontal transmission bias for cooperation, $T_A>T_B$, the cost is not too high $c < (1+b) \frac{T_A-T_B}{1-T_B}$, and the initial frequency is high enough, $\tilde{p}_0 > \tilde{p}^*$, then cooperation will fix in the population,
If there is a horizontal transmission bias for defection, $T_A<T_B$, defection will take over the population.
\end{corollary}

Here, the third equilibrium is
\begin{equation} \label{eq:vert_hori_alpha0_equilibrium}
\tilde{p}^*(\alpha=0) = \frac{c(1-T_B) - (T_A-T_B)}{b (T_A-T_B) },
\end{equation} 
and the cost boundaries are
\begin{equation}\begin{aligned}
\gamma_1(\alpha=0) = -\frac{(T_A - T_B)}{1-T_B}, \quad
\gamma_2(\alpha=0) = \frac{(1+b) (T_A - T_B)}{1-T_B}.
\end{aligned}\end{equation}
If $T_A>T_B$ then $c>0>\gamma_1$ and cooperation cannot increase from rarity, but it can fix if $c<\frac{(1+b) (T_A - T_B)}{1-T_B}$; otherwise defection will fix.

If $T_A<T_B$ then % TODO check this... gamma2<0 means fixation of A???
\\

\begin{corollary}[Complete assortment of transmission and cooperation]
\end{corollary}
In this case $\alpha=1$, and horizontal transmission always occurs together with the cooperative interaction; there is complete assortment.
In this case there are only two equilibria, $\tilde{p}=0$ and $\tilde{p}=1$.
The condition for evolution of cooperation (i.e. global stability of $\tilde{p}=1$) is found by setting $\tilde{p}'>\tilde{p}$, which gives (\autoref{eq:vert_hori_alpha1})
\begin{equation}\label{eq:vert_hori_alpha1}
c < \frac{b \cdot T_A + (T_A - T_B)}{1-T_B}.
\end{equation}
This is equivalent to a result by~\citet[eq.~1]{lewin2017microbes}.

>>>>

Therefore, for cooperation to take over the population (for $\tilde{p}=1$ to be globally stable) we require $\tilde{p}'>\tilde{p}$, that is,
\begin{equation}
  \tilde{p}^2 (1+b-c) + \tilde{p}(1-\tilde{p}) \big[(1-c) (1 - T_B) + (1+b)T_A\big] 
  > \bar{w}\tilde{p} .
\end{equation}
We divide by $\tilde{p}$, set $\bar{w} = 1 + \tilde{p}(b-c)$, and rearrange to get
\begin{equation}
  (1-\tilde{p}) \big[(1-c) (1 - T_B) + (1+b)T_A\big] 
  > 1 -\tilde{p} .
\end{equation}
Dividing by $(1-\tilde{p})$ we find that $\tilde{p}'>\tilde{p}$ if 
\begin{equation} \label{eq:vert_hori_alpha1_condition_proof}
  (1-c) (1 - T_B) + (1+b)T_A
  > 1
\end{equation}
\\

@@

We define the initial frequency as $\tilde{p}_0$ and the cost boundaries
\begin{equation}\begin{aligned}
\gamma_1 = \frac{b \alpha T_A - (T_A - T_B)}{1-T_B}, \quad
\gamma_2 = \frac{b \alpha T_B + (1+b) (T_A - T_B)}{1-T_B}.
\end{aligned}\end{equation}
Applying eqs.~\ref{eq:vert_hori_equilibrium}, \ref{eq:vert_hori_global_condition1}, and \ref{eq:vert_hori_global_condition2} we can summarize the possible outcomes:
%\begin{enumerate} % p* version
%\item Fixation of cooperation,
%	if $T_A>T_B$ and $0<\tilde{p}^*$, or
%	if $T_A>T_B$ and $0<\tilde{p}^*<1$, and $\tilde{p}_0>\tilde{p}^*$, or
%	if $T_A<T_B$ and $1<\tilde{p}^*$.
%\item Global fixation of defection,
%	if $T_A>T_B$ and $\tilde{p}^*>1$, or
%	if $T_A>T_B$ and $0<\tilde{p}^*<1$, and $\tilde{p}_0<\tilde{p}^*$, or
%	if $T_A<T_B$ and $\tilde{p}^*<0$
%\item Co-existence of both phenotypes at $\tilde{p}^*$, 
%	if $T_A<T_B$ and $0<\tilde{p}^*<1$.
%\end{enumerate}
\begin{enumerate} % gamma version
\item Fixation of cooperation,
	if $T_A>T_B$ and $c<\gamma_1$; or
	if $T_A>T_B$ and $\gamma_1<c<\gamma_2$ and $\tilde{p}_0>\tilde{p}^*$; or
	if $T_A<T_B$ and $c<\gamma_2$.
\item Fixation of defection,
	if $T_A>T_B$ and $\gamma_2<c$; or
	if $T_A>T_B$ and $\gamma_1<c<\gamma_2$ and $\tilde{p}_0<\tilde{p}^*$; or
	if $T_A<T_B$ and $\gamma_1<c$.
\item Coexistence of both phenotypes at $\tilde{p}^*$, 
	if $T_A<T_B$ and $\gamma_2<c<\gamma_1$.

Much of the literature on evolution of cooperation focuses on conditions for cooperation to invade a population of defectors.
The next corollary deals with such a condition.
\\
\end{enumerate}

\begin{corollary}[Condition for cooperation to increase from rarity]
If the initial frequency of the cooperative phenotype if very close to zero, $\tilde{p}_0 \approx 0$, then its frequency will increase if 
\begin{equation} \label{eq:unequal_transmission_from_rarity}
\begin{aligned}
T_A>T_B \;\; \text{and} \;\; c < \gamma_1, \quad \text{or} \quad
T_A<T_B \;\; \text{and} \;\; \gamma_2<c < \gamma_1. 
\end{aligned}
\end{equation} 
\end{corollary}

@@

\begin{corollary}[No assortment of transmission and cooperation]
When $\alpha=0$, then either
\textbf{(1)} cooperation can fix, but not increase from rarity, if there is horizontal bias for cooperation ($T_A>T_B$) and the cost is low enough ($c<(1+b)(T_A-T_B)/(1-T_B)$), or 
\textbf{(2)} cooperation can increase from rarity to a stable coexistence with defection, if there is horizontal bias for defection ($T_A<T_B$) and the cost is low enough ($c<(T_B-T_A)/(1-T_B)$).
\end{corollary}

Here, the third equilibrium is
\begin{equation} \label{eq:vert_hori_alpha0_equilibrium}
\tilde{p}^*(\alpha=0) = \frac{c(1-T_B) - (T_A-T_B)}{b (T_A-T_B) },
\end{equation} 
and the cost boundaries are
\begin{equation}\begin{aligned}
\gamma_1(\alpha=0) = \frac{T_B - T_A}{1-T_B}, \quad
\gamma_2(\alpha=0) = \frac{(1+b) (T_A - T_B)}{1-T_B}.
\end{aligned}\end{equation}
If $T_A>T_B$ then $\gamma_1(\alpha=0)<0<\gamma_2(\alpha=0)$.
So cooperation cannot increase from rarity, but it can fix if $c<\gamma_2(\alpha=0)$.
If $T_A<T_B$ then $\gamma_2(\alpha=0)<0<\gamma_1(\alpha=0)$.
So cooperation cannot fix, but if $c<\gamma_1(\alpha=0)$ then it can increase from rarity to a stable coexistence at $\tilde{p}^*(\alpha=0)$.
\\

\begin{corollary}[Complete assortment of transmission and cooperation]
When $\alpha=1$, there are only two equilibria, $\tilde{p}=0$ and $\tilde{p}=1$.
The condition for evolution of cooperation (i.e. global stability of $\tilde{p}=1$) is found by setting $\tilde{p}'>\tilde{p}$, which gives
\begin{equation}\label{eq:vert_hori_alpha1}
c < \frac{b \cdot T_A + (T_A - T_B)}{1-T_B}.
\end{equation}
\end{corollary}
This is proven in \autoref{eq:vert_hori_alpha1_condition_proof}.
In this case there is complete assortment, and horizontal transmission always occurs together with the cooperative interaction. The same occurs in \citet{lewin2017microbes}, and therefore this corollary is equivalent to their result, see their eq.~1.

In terms of the cost boundaries, \autoref{eq:vert_hori_alpha1} is equivalent to $c<\gamma_1$. If $T_A>T_B$ then that suffices for fixation of cooperation. If $T_B>T_A$ then $\gamma_2(\alpha=1)<0$ and again, \autoref{eq:vert_hori_alpha1} is sufficient for increase in frequency of $A$ up to $\tilde{p}^*(\alpha=1) \approx \infty$.
